---
title: "PLAD ADM-2 Panel — Construction Plan"
subtitle: "Balanced ADM-2 × year panel, 2000–2022"
author: "CERDI, Université Clermont-Auvergne"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    df-print: kable
execute:
  warning: false
  message: false
knitr:
  opts_knit:
    root.dir: "../"
---

## Overview

This document constructs a balanced ADM-2 × year panel for 2000–2022 matching political
leaders' birthplaces and ethnicities (PLAD v7.0) to subnational administrative units
(GADM 4.10).

**Output:** `output/plad_adm2_panel.rds` — 48,010 ADM-2 units × 23 years = 1,104,230 rows

| Variable | Definition |
|----------|-----------|
| `birthplace_match` | 1 if ADM-2 contains the current leader's birthplace in year *t* |
| `ethnicity_match` | 1 if ADM-2 intersects the GeoEPR settlement area of the leader's ethnicity in year *t* |

---

## Setup

```{r setup}
library(sf)
library(haven)
library(tidyverse)
library(countrycode)

sf_use_s2(FALSE)  # GADM geometries are invalid under s2 spherical rules; planar GEOS is accurate enough for ADM-2 analysis

stopifnot(
  file.exists("source/gadm_410_adm2_clean.gpkg"),
  file.exists("source/PLAD/PLAD_April_2024.dta"),
  file.exists("source/GeoEPR/GeoEPR-2023.geojson")
)
```

---

## 1. Load Data

### 1.1 GADM ADM-2 boundaries

```{r load-gadm}
gadm <- st_read("source/gadm_410_adm2_clean.gpkg", quiet = TRUE) |>
  select(gid_2, name_adm2, name_adm1, iso3, country, geom)

gadm_df <- st_drop_geometry(gadm)

cat("GADM: ", nrow(gadm), "ADM-2 units across", n_distinct(gadm$iso3), "countries\n")
```

### 1.2 PLAD leaders

```{r load-plad}
plad_raw <- read_dta("source/PLAD/PLAD_April_2024.dta") |>
  zap_labels() |>
  select(gid_0, gid_2, startyear, endyear, latitude, longitude,
         geo_precision, ethnicity_geoepr1) |>
  mutate(
    gid_0             = as.character(gid_0),
    gid_2             = na_if(as.character(gid_2), "."),
    startyear         = as.integer(startyear),
    endyear           = as.integer(endyear),
    latitude          = as.numeric(latitude),
    longitude         = as.numeric(longitude),
    geo_precision     = as.integer(geo_precision),
    ethnicity_geoepr1 = na_if(as.character(ethnicity_geoepr1), "")
  )

cat("PLAD: ", nrow(plad_raw), "leaders\n")
cat("Missing gid_2:", sum(is.na(plad_raw$gid_2)), "\n")
cat("geo_precision distribution:\n")
print(table(plad_raw$geo_precision, useNA = "always"))
```

### 1.3 GeoEPR ethnic settlement areas

```{r load-geoepr}
geoepr <- st_read("source/GeoEPR/GeoEPR-2023.geojson", quiet = TRUE) |>
  select(group, gwid, from, to, geometry) |>
  mutate(
    iso3 = countrycode(as.integer(gwid), "gwn", "iso3c"),
    from = as.integer(from),
    to   = as.integer(to)
  )

cat("GeoEPR: ", nrow(geoepr), "group-period polygons,", n_distinct(geoepr$group), "unique groups\n")
```

---

## 2. Expand PLAD to Leader × Year

Each leader row spans `startyear`–`endyear`. This step creates one row per leader per
calendar year, clipped to 2000–2022.

```{r expand-plad}
plad <- plad_raw |>
  rename(iso3 = gid_0) |>
  filter(!is.na(iso3), !is.na(startyear), !is.na(endyear),
         startyear <= 2022L, endyear >= 2000L) |>
  mutate(
    yr_start = pmax(startyear, 2000L),
    yr_end   = pmin(endyear,   2022L)
  ) |>
  rowwise() |>
  mutate(year = list(seq(yr_start, yr_end))) |>
  ungroup() |>
  unnest(year) |>
  select(-startyear, -endyear, -yr_start, -yr_end)

# Guard: one effective leader per country-year
dups <- count(plad, iso3, year) |> filter(n > 1)
if (nrow(dups) > 0) {
  warning(sprintf("%d iso3-year pairs have multiple leaders; keeping first.", nrow(dups)))
  plad <- group_by(plad, iso3, year) |> slice(1) |> ungroup()
}

cat("Leader-year rows: ", nrow(plad), "\n")
cat("Countries covered: ", n_distinct(plad$iso3), "\n")
cat("Year range: ", range(plad$year), "\n")
```

---

## 3. Panel Spine

All 48,010 GADM ADM-2 units × 23 years (2000–2022) = 1,104,230 rows.
Leader info is joined by `(iso3, year)`. ADM-2 units in non-PLAD countries receive `NA`.

```{r spine}
years <- 2000L:2022L

spine <- crossing(
  gadm_df |> select(gid_2, name_adm2, name_adm1, iso3, country),
  year = years
) |>
  left_join(
    plad |> select(iso3, year,
                   gid_2_ldr = gid_2, latitude, longitude,
                   geo_precision, ethnicity_geoepr1),
    by = c("iso3", "year")
  )

cat("Spine rows: ", nrow(spine), "\n")
cat("Rows with a matched leader: ", sum(!is.na(spine$gid_2_ldr) | !is.na(spine$ethnicity_geoepr1)), "\n")
```

---

## 4. `birthplace_match`

### Logic

| Case | Condition | Result |
|------|-----------|--------|
| Primary | `gid_2` present in PLAD | Exact match: `gid_2 == gid_2_ldr` |
| Fallback | `gid_2` missing + `geo_precision < 4` | `st_within(lat/lon point, ADM-2 polygon)` |
| NA | `gid_2` missing + `geo_precision ≥ 4` | `NA` — location too coarse for ADM-2 |
| No leader | Country not in PLAD | `NA` |

```{r birthplace-primary}
# Primary: exact gid_2 match
spine <- spine |>
  mutate(
    birthplace_match = case_when(
      !is.na(gid_2_ldr) ~ as.integer(gid_2 == gid_2_ldr),
      TRUE               ~ NA_integer_
    )
  )
```

```{r birthplace-fallback}
# Spatial fallback: missing gid_2 + geo_precision < 4
fallback <- plad |>
  filter(is.na(gid_2), !is.na(latitude), !is.na(longitude),
         !is.na(geo_precision), geo_precision < 4L) |>
  distinct(iso3, year, latitude, longitude)

cat("Leader-years eligible for spatial fallback:", nrow(fallback), "\n")

if (nrow(fallback) > 0) {
  pts      <- st_as_sf(fallback, coords = c("longitude", "latitude"), crs = 4326)
  within_i <- st_within(pts, gadm, sparse = TRUE)
  fallback$gid_2_fallback <- vapply(
    within_i,
    function(i) if (length(i) == 0L) NA_character_ else gadm$gid_2[i[1L]],
    character(1)
  )

  cat("Fallback resolved: ", sum(!is.na(fallback$gid_2_fallback)), "/", nrow(fallback), "\n")

  spine <- spine |>
    left_join(fallback |> select(iso3, year, gid_2_fallback), by = c("iso3", "year")) |>
    mutate(
      birthplace_match = case_when(
        !is.na(birthplace_match) ~ birthplace_match,
        !is.na(gid_2_fallback)   ~ as.integer(gid_2 == gid_2_fallback),
        TRUE                     ~ NA_integer_
      )
    ) |>
    select(-gid_2_fallback)
}
```

```{r birthplace-summary}
cat("birthplace_match distribution:\n")
print(table(spine$birthplace_match, useNA = "always"))
```

---

## 5. `ethnicity_match`

For each (leader, year), find the GeoEPR polygon for `ethnicity_geoepr1` in that year,
then mark all ADM-2 units that spatially intersect it.

Country link: PLAD `iso3` (= `gid_0`) ↔ GeoEPR `iso3` (derived from `gwid` via `countrycode`).

```{r ethnicity-match}
gadm_by_iso3 <- split(gadm, gadm$iso3)

eth_leaders <- plad |>
  filter(!is.na(ethnicity_geoepr1)) |>
  distinct(iso3, year, ethnicity_geoepr1)

cat("Unique (iso3, year, ethnicity) combinations:", nrow(eth_leaders), "\n")

eth_list <- vector("list", nrow(eth_leaders))

for (k in seq_len(nrow(eth_leaders))) {
  ldr      <- eth_leaders[k, ]
  grp_geom <- geoepr |>
    filter(iso3 == ldr$iso3,
           group == ldr$ethnicity_geoepr1,
           from <= ldr$year, to >= ldr$year)
  gadm_cty <- gadm_by_iso3[[ldr$iso3]]

  if (nrow(grp_geom) == 0L || is.null(gadm_cty) || nrow(gadm_cty) == 0L) next

  hits <- st_intersects(gadm_cty, st_union(grp_geom), sparse = TRUE)
  eth_list[[k]] <- tibble(
    gid_2           = gadm_cty$gid_2,
    year            = ldr$year,
    ethnicity_match = as.integer(lengths(hits) > 0L)
  )
}

eth_df <- bind_rows(eth_list)

if (nrow(eth_df) > 0) {
  eth_df <- eth_df |>
    group_by(gid_2, year) |>
    summarise(ethnicity_match = as.integer(any(ethnicity_match == 1L)), .groups = "drop")
}

n_unmatched <- sum(vapply(eth_list, is.null, logical(1)))
cat("Leader-years with no GeoEPR match:", n_unmatched, "/", nrow(eth_leaders), "\n")
cat("(Missing polygon or group name mismatch)\n")

spine <- left_join(spine, eth_df, by = c("gid_2", "year"))

cat("\nethnicity_match distribution:\n")
print(table(spine$ethnicity_match, useNA = "always"))
```

---

## 6. Assemble and Save

```{r assemble}
panel <- spine |>
  select(gid_2, name_adm2, name_adm1, iso3, country, year,
         birthplace_match, ethnicity_match) |>
  arrange(iso3, gid_2, year)

saveRDS(panel, "output/plad_adm2_panel.rds")

cat("Saved: output/plad_adm2_panel.rds\n")
cat("Dimensions:", nrow(panel), "x", ncol(panel), "\n")
```

---

## 7. Verification

```{r verify}
stopifnot(file.exists("output/plad_adm2_panel.rds"))

p <- readRDS("output/plad_adm2_panel.rds")

# Dimension check
cat("Expected: 1,104,230 rows. Actual:", nrow(p), "\n")
cat("Expected: 8 columns. Actual:    ", ncol(p), "\n")
stopifnot(nrow(p) == nrow(gadm_df) * length(years))

# Variable ranges
cat("\nbirthplace_match in {0, 1, NA}:",
    all(p$birthplace_match %in% c(0L, 1L, NA_integer_)), "\n")
cat("ethnicity_match  in {0, 1, NA}:",
    all(p$ethnicity_match  %in% c(0L, 1L, NA_integer_)), "\n")

# Plausibility: at least one birthplace match per leader-year
n_bp_matches <- sum(p$birthplace_match == 1L, na.rm = TRUE)
cat("\nTotal birthplace_match == 1 cells:", n_bp_matches, "\n")
cat("Total ethnicity_match  == 1 cells:", sum(p$ethnicity_match == 1L, na.rm = TRUE), "\n")
stopifnot(n_bp_matches > 0)

cat("\nAll checks passed.\n")
```
