---
title: "PLAD ADM-2 Cross-section — Construction Plan"
subtitle: "Balanced ADM-2 cross-section, 2000–2022 summary"
author: "CERDI, Université Clermont-Auvergne"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: true
    df-print: kable
execute:
  warning: false
  message: false
knitr:
  opts_knit:
    root.dir: "../"
---

## Overview

This document collapses the balanced ADM-2 × year panel (`output/plad_adm2_panel.rds`)
into a time-invariant cross-section summarising leader birthplace affiliation over 2000–2022.

**Dependency:** requires `script/plad_panel.qmd` to have been rendered first.

**Output:** `output/plad_adm2_crosssection.rds` — 48,010 rows (one per ADM-2 unit)

| Variable | Definition |
|----------|-----------|
| `ever_birthplace` | 1 if ADM-2 was ever a leader's birthplace during 2000–2022 |
| `always_birthplace` | 1 if ADM-2 was a leader's birthplace in every non-NA year of 2000–2022 |
| `pct_birthplace` | Share of non-NA years in which ADM-2 was a leader's birthplace |

---

## Setup

```{r setup}
library(tidyverse)

if (!file.exists("output/plad_adm2_panel.rds")) {
  stop("Panel output not found. Render script/plad_panel.qmd first.")
}
```

---

## 1. Load Panel

```{r load-panel}
panel <- readRDS("output/plad_adm2_panel.rds")

cat("Panel dimensions:", nrow(panel), "x", ncol(panel), "\n")
cat("ADM-2 units:     ", n_distinct(panel$gid_2), "\n")
cat("Years covered:   ", paste(range(panel$year), collapse = "–"), "\n")
```

---

## 2. Collapse to ADM-2 Level

Group by ADM-2 unit and compute the three birthplace summary variables.

**NA handling:** if all 23 years have `birthplace_match = NA` (country not in PLAD),
all three variables are set to `NA` rather than `0`/`1`.

```{r crosssection}
# Metadata: one row per gid_2 (take first occurrence to handle any inconsistency)
gadm_meta <- panel |>
  distinct(gid_2, name_adm2, name_adm1, iso3, country) |>
  group_by(gid_2) |>
  slice(1) |>
  ungroup()

# Aggregate birthplace variables by gid_2 only
crosssection <- panel |>
  group_by(gid_2) |>
  summarise(
    n_non_na          = sum(!is.na(birthplace_match)),
    ever_birthplace   = if_else(
      n_non_na == 0L,
      NA_integer_,
      as.integer(any(birthplace_match == 1L, na.rm = TRUE))
    ),
    always_birthplace = if_else(
      n_non_na == 0L,
      NA_integer_,
      as.integer(all(birthplace_match == 1L, na.rm = TRUE))
    ),
    pct_birthplace    = if_else(
      n_non_na == 0L,
      NA_real_,
      mean(birthplace_match, na.rm = TRUE)
    ),
    .groups = "drop"
  ) |>
  select(-n_non_na) |>
  left_join(gadm_meta, by = "gid_2") |>
  select(gid_2, name_adm2, name_adm1, iso3, country,
         ever_birthplace, always_birthplace, pct_birthplace)

cat("Cross-section dimensions:", nrow(crosssection), "x", ncol(crosssection), "\n")
```

---

## 3. Summary Statistics

```{r summary}
cat("ever_birthplace:\n")
print(table(crosssection$ever_birthplace, useNA = "always"))

cat("\nalways_birthplace:\n")
print(table(crosssection$always_birthplace, useNA = "always"))

cat("\npct_birthplace (non-NA units):\n")
summary(crosssection$pct_birthplace[!is.na(crosssection$pct_birthplace)])
```

ADM-2 units with `ever_birthplace == 1`:

```{r ever-table}
crosssection |>
  filter(ever_birthplace == 1L) |>
  select(iso3, name_adm1, name_adm2, pct_birthplace) |>
  arrange(iso3, desc(pct_birthplace))
```

---

## 4. Save

```{r save}
saveRDS(crosssection, "output/plad_adm2_crosssection.rds")
cat("Saved: output/plad_adm2_crosssection.rds\n")
```

---

## 5. Verification

```{r verify}
stopifnot(file.exists("output/plad_adm2_crosssection.rds"))

cs <- readRDS("output/plad_adm2_crosssection.rds")

# Dimension check
cat("Rows:   ", nrow(cs), "(unique gid_2:", n_distinct(cs$gid_2), ")\n")
cat("Columns:", ncol(cs), "\n")
stopifnot(n_distinct(cs$gid_2) == nrow(cs))  # one row per ADM-2 unit

# Variable range checks
cat("\never_birthplace   in {0, 1, NA}:",
    all(cs$ever_birthplace   %in% c(0L, 1L, NA_integer_)), "\n")
cat("always_birthplace in {0, 1, NA}:",
    all(cs$always_birthplace %in% c(0L, 1L, NA_integer_)), "\n")
cat("pct_birthplace    in [0, 1]:    ",
    all(cs$pct_birthplace >= 0 & cs$pct_birthplace <= 1, na.rm = TRUE), "\n")

# Consistency: always_birthplace <= ever_birthplace
cat("always <= ever (non-NA):        ",
    all(cs$always_birthplace <= cs$ever_birthplace, na.rm = TRUE), "\n")

cat("\nAll checks passed.\n")
```
